<<<<<<< HEAD
---
title: "Lab5"
author: "Patrick Casanas"
date: "2024-10-03"
output: github_document
---

```{r}
library(data.table)
library(dtplyr)
library (dplyr)
library(leaflet)
stations <- fread("https://noaa-isd-pds.s3.amazonaws.com/isd-history.csv")
stations[, USAF := as.integer(USAF)]

stations[, USAF   := fifelse(USAF == 999999, NA_integer_, USAF)]
stations[, CTRY   := fifelse(CTRY == "", NA_character_, CTRY)]
stations[, STATE  := fifelse(STATE == "", NA_character_, STATE)]

stations <- unique(stations[, list(USAF, CTRY, STATE)])

stations <- stations[!is.na(USAF)]

stations[, n := 1:.N, by = .(USAF)]
stations <- stations[n == 1,][, n := NULL]

met <- fread("https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz")
```

```{r}
met_merged <- merge(
  x=met,
  y=stations,
  by.x="USAFID",
  by.y="USAF",
  all.x=TRUE,
  all.y=FALSE
)
head (met_merged)
```

```{r, eval=TRUE}

median_temp <- median(met_merged$temp, na.rm = TRUE)
median_wind <- median(met_merged$wind.sp, na.rm = TRUE)
median_pressure <- median(met_merged$atm.press, na.rm = TRUE)

closest_temp_station <- met_merged[!is.na(met_merged$temp)][which.min(abs(met_merged$temp - median_temp)), ]
closest_wind_station <- met_merged[!is.na(met_merged$wind.sp)][which.min(abs(met_merged$wind.sp - median_wind)), ]
closest_pressure_station <- met_merged[!is.na(met_merged$atm.press)][which.min(abs(met_merged$atm.press - median_pressure)), ]

print("Station closest to median temperature:")
print(closest_temp_station)

print("Station closest to median wind speed:")
print(closest_wind_station)

print("Station closest to median atmospheric pressure:")
print(closest_pressure_station)
```

The closest station to the median temperature is station 720113.

The closest station to the median wind speed is station 690150.

The closest station to the median atmospheric pressure is also 690150.

**Question 2**

```{r, eval=TRUE}

representative_stations <- met_merged %>%
  filter(!is.na(temp), !is.na(wind.sp), !is.na(atm.press), !is.na(elev)) %>%
  group_by(STATE) %>%
  mutate(
    med_temp = median(temp),
    med_wind = median(wind.sp),
    med_pressure = median(atm.press),
    euclidean_dist = sqrt((temp - med_temp)^2 + (wind.sp - med_wind)^2 + (atm.press - med_pressure)^2)
  ) %>%
  filter(euclidean_dist == 0) %>%
  arrange(STATE, elev) %>%
  slice(1)

print(met_merged)

```

**Question 3**

```{r, eval=TRUE}

midpoint_stations <- met_merged %>%
  group_by(STATE) %>%
  summarize(
    mid_lat = mean(lat, na.rm = TRUE),
    mid_lon = mean(lon, na.rm = TRUE)
  )

print(midpoint_stations)

met_with_midpoints <- merge(met_merged, midpoint_stations, by = "STATE")

met_with_midpoints <- met_with_midpoints %>%
  mutate(
    dist_to_mid = sqrt((lat - mid_lat)^2 + (lon - mid_lon)^2)  # Use lowercase 'lat' and 'lon'
  )

print(head(met_with_midpoints))

closest_midpoint_station <- met_with_midpoints %>%
  group_by(STATE) %>%
  filter(dist_to_mid == min(dist_to_mid)) %>%
  slice(1)

print(closest_midpoint_station)

combined_stations <- bind_rows(closest_midpoint_station, representative_stations)

print(combined_stations)

```

```{r, eval=FALSE}
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = combined_stations,
                   lng = ~lon, lat = ~lat)
```

**Question 4**

```{r, eval=TRUE}

state_avg_temp <- met_merged %>%
  group_by(STATE) %>%
  summarize(avg_temp = mean(temp, na.rm = TRUE))

state_avg_temp <- state_avg_temp %>%
  mutate(
    temp_level = ifelse(avg_temp < 20, "Low", 
                  ifelse(avg_temp >= 20 & avg_temp < 25, "Mid", "High"))
  )

met_with_levels <- merge(met_merged, state_avg_temp, by = "STATE")

print(head(met_with_levels))

summary_table <- met_with_levels %>%
  group_by(temp_level) %>%
  summarize(
    num_records = n(),
    num_na_temp = sum(is.na(temp)),
    num_na_wind = sum(is.na(wind.sp)),
    num_na_pressure = sum(is.na(atm.press)),
    num_stations = n_distinct(USAFID),
    num_states = n_distinct(STATE),
    mean_temp = mean(temp, na.rm = TRUE),
    mean_wind = mean(wind.sp, na.rm = TRUE),
    mean_pressure = mean(atm.press, na.rm = TRUE)
  )

print(summary_table)
```
=======
---
title: "Lab5"
author: "Patrick Casanas"
date: "2024-10-03"
output: github_document
---

```{r}
library(data.table)
library(dtplyr)
library (dplyr)
library(leaflet)
stations <- fread("https://noaa-isd-pds.s3.amazonaws.com/isd-history.csv")
stations[, USAF := as.integer(USAF)]

stations[, USAF   := fifelse(USAF == 999999, NA_integer_, USAF)]
stations[, CTRY   := fifelse(CTRY == "", NA_character_, CTRY)]
stations[, STATE  := fifelse(STATE == "", NA_character_, STATE)]

stations <- unique(stations[, list(USAF, CTRY, STATE)])

stations <- stations[!is.na(USAF)]

stations[, n := 1:.N, by = .(USAF)]
stations <- stations[n == 1,][, n := NULL]

met <- fread("https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz")
```

```{r}
met_merged <- merge(
  x=met,
  y=stations,
  by.x="USAFID",
  by.y="USAF",
  all.x=TRUE,
  all.y=FALSE
)
head (met_merged)
```

```{r, eval=TRUE}

median_temp <- median(met_merged$temp, na.rm = TRUE)
median_wind <- median(met_merged$wind.sp, na.rm = TRUE)
median_pressure <- median(met_merged$atm.press, na.rm = TRUE)

closest_temp_station <- met_merged[!is.na(met_merged$temp)][which.min(abs(met_merged$temp - median_temp)), ]
closest_wind_station <- met_merged[!is.na(met_merged$wind.sp)][which.min(abs(met_merged$wind.sp - median_wind)), ]
closest_pressure_station <- met_merged[!is.na(met_merged$atm.press)][which.min(abs(met_merged$atm.press - median_pressure)), ]

print("Station closest to median temperature:")
print(closest_temp_station)

print("Station closest to median wind speed:")
print(closest_wind_station)

print("Station closest to median atmospheric pressure:")
print(closest_pressure_station)
```

The closest station to the median temperature is station 720113.

The closest station to the median wind speed is station 690150.

The closest station to the median atmospheric pressure is also 690150.

**Question 2**

```{r, eval=TRUE}

representative_stations <- met_merged %>%
  filter(!is.na(temp), !is.na(wind.sp), !is.na(atm.press), !is.na(elev)) %>%
  group_by(STATE) %>%
  mutate(
    med_temp = median(temp),
    med_wind = median(wind.sp),
    med_pressure = median(atm.press),
    euclidean_dist = sqrt((temp - med_temp)^2 + (wind.sp - med_wind)^2 + (atm.press - med_pressure)^2)
  ) %>%
  filter(euclidean_dist == 0) %>%
  arrange(STATE, elev) %>%
  slice(1)

print(met_merged)

```

**Question 3**

```{r, eval=TRUE}

midpoint_stations <- met_merged %>%
  group_by(STATE) %>%
  summarize(
    mid_lat = mean(lat, na.rm = TRUE),
    mid_lon = mean(lon, na.rm = TRUE)
  )

print(midpoint_stations)

met_with_midpoints <- merge(met_merged, midpoint_stations, by = "STATE")

met_with_midpoints <- met_with_midpoints %>%
  mutate(
    dist_to_mid = sqrt((lat - mid_lat)^2 + (lon - mid_lon)^2)  # Use lowercase 'lat' and 'lon'
  )

print(head(met_with_midpoints))

closest_midpoint_station <- met_with_midpoints %>%
  group_by(STATE) %>%
  filter(dist_to_mid == min(dist_to_mid)) %>%
  slice(1)

print(closest_midpoint_station)

combined_stations <- bind_rows(closest_midpoint_station, representative_stations)

print(combined_stations)

```

```{r, eval=FALSE}
leaflet() %>%
  addTiles() %>%
  addCircleMarkers(data = combined_stations,
                   lng = ~lon, lat = ~lat)
```
>>>>>>> 700af6f74f78a9a031cf6b4af4ee396b9a730b45
